---
title: "MB5370 Module 4 - QFish Report"
author: "Tory Pritchard"
date: "2024-05-16"
output: html_document
---

#____________________________________________________________________________

Workspace/Rmd file location
C:/Users/Admin/OneDrive/Documents/MB5370/Module_04/Mod4_Personal/QFISH_report.Rmd
File also in Git hub repo!

Description: This assignment uses real-world marine data collected as part of Queensland fisheries QFISH database to develop a report and produce a plot visualizing the data.

#____________________________________________________________________________

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

```{r}
#Load required packages
library(DataViz)
library(palmerpenguins)
library(ggthemes)
library(wesanderson)
library(readr)
library(tidyverse)
```



### Part 1: Obtain data

Data was obtained from the Department of Agriculture and Fisheries website under QFish. The chosen data set was:
[Commercial Line Fishery](https://qfish.fisheries.qld.gov.au/query/commercial-line-fishery/table)

Before downloading the data, in the `Customize` tab, the `FishingMethod` was moved to `Row Area` to remove the data from the `Filter Area`. 

Data was exported as Comma Separated Values (CSV) and available in Github repo:  [Mod4_Personal](https://github.com/tkpritch/Mod4_Personal.git)



### Part 2: Tidy data

Import dataset from Github.
```{r}
fishmethod <- read_csv("../data/CommLineFish.csv")
```
Remove/filter out the rows of Area that include more than just the numerical year value
Change "N/A" values to NA.
```{r}
filtered_fishmethod <- fishmethod %>%
  filter(!grepl("Total", CalendarYear)) %>%
  filter(CalendarYear != "2024 incomplete") %>%
  mutate_all(~ ifelse(. == "N/A", NA, .))
```

Remove the commas in the Tonnes column so it can be identified as a numeric type.
```{r}
filtered_fishmethod$Tonnes <- as.numeric(gsub("[,]", "", filtered_fishmethod$Tonnes))
```

Assign the proper variable type to the rows (i.e. numeric).
```{r}
filtered_fishmethod <- filtered_fishmethod %>%
  mutate(Tonnes = as.numeric(Tonnes),
         CalendarYear = as.numeric(CalendarYear))
```

Exclude the fishing methods that are all or almost all NA.
```{r}
#List methods to be excluded
methods_to_exclude <- c("Trap", "Other", "Diving", "Danish Seine", "Collection")

#Filter out methods
filtered_method <- filtered_fishmethod %>%
  filter(!FishingMethod %in% methods_to_exclude)
```
Now data should be ready to plot!




### Part 3: Plot 

Data included catch tonnage in the commercial line fishery, but other methods were included for comparison of how much was caught among different methods from the years 1990 to 2023.

Plot data showing tonnes caught over the years by each method.
```{r}
#| fig.cap = "Figure 3. Commercial line fishery catch data for various species from 1990 to 2023. Fishing methods plotted are listed. Points that do not connect indicate there was no data recorded for those years. Fishing methods that were excluded from the plot due to having no data recorded include: Trap, Other, Diving, Danish Seine, Collection. Data obtained from Department of Agriculture and Fisheries, QFish."
ggplot(filtered_method, aes(x = CalendarYear, y = Tonnes, colour = FishingMethod)) +
  geom_line() +
  geom_point() +
  scale_color_manual(values = palette.colors(palette = "Dark 2")) +
  labs(x= "Year", colour = "Method of Capture")
```
