---
title: "MB5370 Module 4 - QFISH_report"
author: "Tory Pritchard"
date: "2024-05-16"
output: html_document
---

#____________________________________________________________________________

Workspace/Rmd file location
C:/Users/Admin/OneDrive/Documents/MB5370/Module_04/Mod4_Personal/QFISH_report.Rmd
File also in Git hub!

Description: This assignment uses real-world marine data collected as part of Queensland fisheries QFISH database to develop a report.

#____________________________________________________________________________

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

```{r}
#Load required packages
library(DataViz)
library(palmerpenguins)
library(ggthemes)
library(readr)
library(tidyverse)
```


### Part 1: Obtain data

Data was obtained from the Department of Agriculture and Fisheries website under QFish. The chosen data set was:
[Commercial Line Fishery](https://qfish.fisheries.qld.gov.au/query/commercial-line-fishery/table)

Before downloading the data, in the `Customize` tab, the `FishingMethod` was moved to `Row Area` to remove the data from the `Filter Area`. 

Data was exported as Comma Separated Values (CSV) and available in Github repo:  [Mod4_Personal](https://github.com/tkpritch/Mod4_Personal.git)


### Part 2: Tidy data

Import dataset from Github.
```{r}
fishmethod <- read_csv("../data/CommLineFish.csv")
```
Remove/filter out the rows of Area that include more than just the year, and change "N/A" values to NA.
```{r}
filtered_fishmethod <- fishmethod %>%
  filter(!grepl("Total", CalendarYear)) %>%
  filter(CalendarYear != "2024 incomplete") %>%
  mutate_all(~ ifelse(. == "N/A", NA, .))
```

Remove the commas in the Tonnes column so it can be identified as a numeric type.
```{r}
filtered_fishmethod$Tonnes <- as.numeric(gsub("[,]", "", filtered_fishmethod$Tonnes))
```

Since the columns of data had character and numeric values in the cells, mutate the data to assign the proper variable type to the rows.
```{r}
filtered_fishmethod <- filtered_fishmethod %>%
  mutate(Tonnes = as.numeric(Tonnes),
         CalendarYear = as.numeric(CalendarYear))
```

Exclude the fishing methods that are all or almost all 0's.
```{r}
#List methods to be excluded
methods_to_exclude <- c("Trap", "Other", "Diving", "Danish Seine", "Collection")

#Filter out methods
filtered_method <- filtered_fishmethod %>%
  filter(!FishingMethod %in% methods_to_exclude)
```
Now data should be ready to plot!


### Part 3: Plot 

Data included tonnes caught in the commercial line fishery, but other methods were included for comparison of how much is caught amongst different methods from the years 1990 to 2023.

Plot data showing tonnes caught over the years by each method.
```{r}
#| fig.cap = "Figure 3. Commercial line fishery catch data for various species from 1990 to 2023. Fishing methods plotted are listed. Data obtained from Department of Agriculture and Fisheries, QFish."
ggplot(filtered_method, aes(x = CalendarYear, y = Tonnes, colour = FishingMethod),
       labs(x="Year")) +
  geom_point()
  geom_smooth() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  #scale_y_discrete(breaks = seq(0, 15000, by = 2500))

```
