---
title: "MB5370 Module 4 - Workshop3 - Data Wrangling in R"
author: "Tory Pritchard"
date: "2024-05-16"
output: html_document
---

#____________________________________________________________________________

Workspace/Rmd file location
C:/Users/Admin/OneDrive/Documents/MB5370/Module_04/Mod4_Personal/Workshop3.Rmd
File also in Git hub!

Description: This workshop follows the data tidying section of the [R4DS textbook](https://r4ds.hadley.nz/data-tidy#sec-tidy-data). Aim is to take a tabular dataset and prepare it in any way you need.
Data tidying - aka data wrangling - is crucial to actually order to actually use your valuable data.

Tibbles
Used in place of the traditional data frames when using the tidyr package. They are slightly adjusted data frames designed to keep up without breaking existing code. 'Future proof' data frames.

#____________________________________________________________________________

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Install and load tidyverse packages
```{r}
## install.packages("tidyverse") # Delete this line once installed
library("tidyverse")

#and other packages
library(DataViz)
library(ggplot2)
library(tidyr)
library(tidyverse)
library(palmerpenguins)
library(ggthemes)
#devtools::install_github("kjhealy/socviz")
```

## Tidy data
Rearranging data to a format more appropriate for analysis.
See tibble for the first time:
```{r}
table1
#> # A tibble: 6 × 4
#>   country      year  cases population
#>   <chr>       <int>  <int>      <int>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
table2
#> # A tibble: 12 × 4
#>   country      year type           count
#>   <chr>       <int> <chr>          <int>
#> 1 Afghanistan  1999 cases            745
#> 2 Afghanistan  1999 population  19987071
#> 3 Afghanistan  2000 cases           2666
#> 4 Afghanistan  2000 population  20595360
#> 5 Brazil       1999 cases          37737
#> 6 Brazil       1999 population 172006362
#> # ... with 6 more rows
table3
#> # A tibble: 6 × 3
#>   country      year rate             
#> * <chr>       <int> <chr>            
#> 1 Afghanistan  1999 745/19987071     
#> 2 Afghanistan  2000 2666/20595360    
#> 3 Brazil       1999 37737/172006362  
#> 4 Brazil       2000 80488/174504898  
#> 5 China        1999 212258/1272915272
#> 6 China        2000 213766/1280428583
```

How we make our dataset tidy is by following three interrelated rules. 
Each variable must have its own column.
Each observation must have its own row.
Each value must have its own cell.

Picture from textbook can help visualize these rules.

# Pipes intro
A pipe is really only designed to help you better understand what the code is doing.
ex. You take the data, use a pipe and apply a function to it, specifying arguments inside the function (like below, we apply the function mutate to compute the rate given two variables).
```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)
#> # A tibble: 6 × 5
#>   country      year  cases population  rate
#>   <chr>       <int>  <int>      <int> <dbl>
#> 1 Afghanistan  1999    745   19987071 0.373
#> 2 Afghanistan  2000   2666   20595360 1.29 
#> 3 Brazil       1999  37737  172006362 2.19 
#> 4 Brazil       2000  80488  174504898 5.61 
#> 5 China        1999 212258 1272915272 1.67 
#> 6 China        2000 213766 1280428583 1.67

# Compute cases per year
table1 %>% 
  count(year, wt = cases)
#> # A tibble: 2 × 2
#>    year      n
#>   <int>  <int>
#> 1  1999 250740
#> 2  2000 296920

# Visualise changes over time
library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```

Understanding whether your data frame structure is optimal (or tidy) is a fundamental skill for a data scientist.

EXERCISE to do later!!!!
For each of the sample tables, describe what each observation and each column represents.
Sketch out the processes you would use to calculate the rate for table2 and table3. You will need to perform four operations:
Extract the number of TB cases per country per year
Extract the matching population per country per year
Divide cases by population, and multiply by 10,000
Store back in the appropriate place
Hint: you haven’t yet learned the functions you need to actually perform these, but you can still think through the transformations!

```{r}

```

## Pivoting data to it tidy
First step in tidying is to understand what each variable and observation means.
Second step is to resolve one of the two common problems with untidy. These are:
  One variable is spread across multiple columns
  One observation is scattered across multiple rows

To fix these we will pivot our data (i.e. move it around) into tidy form using two functions in tidyr: pivot_longer() to lengthen data and pivot_wider() to widen data.

### Lengthening datasets
Make out data frame longer with `pivot_longer()`
In this dataset, each observation is a song.
```{r}
billboard
#> # A tibble: 317 × 79
#>   artist       track               date.entered   wk1   wk2   wk3   wk4   wk5
#>   <chr>        <chr>               <date>       <dbl> <dbl> <dbl> <dbl> <dbl>
#> 1 2 Pac        Baby Don't Cry (Ke... 2000-02-26      87    82    72    77    87
#> 2 2Ge+her      The Hardest Part O... 2000-09-02      91    87    92    NA    NA
#> 3 3 Doors Down Kryptonite          2000-04-08      81    70    68    67    66
#> 4 3 Doors Down Loser               2000-10-21      76    76    72    69    67
#> 5 504 Boyz     Wobble Wobble       2000-04-15      57    34    25    17    17
#> 6 98^0         Give Me Just One N... 2000-08-19      51    39    34    26    26
#> # ℹ 311 more rows
#> # ℹ 71 more variables: wk6 <dbl>, wk7 <dbl>, wk8 <dbl>, wk9 <dbl>, ...
```

We need to have the data in a format where each row is an observation. Long format. 
```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank"
  )
#> # A tibble: 24,092 × 5
#>    artist track                   date.entered week   rank
#>    <chr>  <chr>                   <date>       <chr> <dbl>
#>  1 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk1      87
#>  2 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk2      82
#>  3 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk3      72
#>  4 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk4      77
#>  5 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk5      87
#>  6 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk6      94
#>  #> # ℹ 24,082 more rows
```

There are three key arguments to the pivot_longer() function:
  cols which specifies the columns you want to pivot (the ones that aren’t variables). Note: you could either use !c(artist, track, date.entered) OR starts_with('wk') because the cols argument uses the same syntax as select().
  names_to which names the variable stored in the column names. We chose to name that variable week.
  values_to which names the variable stored in the cell values that we named rank
  
Note: in the code "week" and "rank" are quoted because they are new variables that we are creating, they don’t exist yet in the data when we run the pivot_longer() call.

Notice the NA values in the output above? It looks like “Baby Don’t Cry” by 2 Pac was only in the top 100 for 7 out of 76 weeks. Therefore, when we lengthened the data, the weeks where it wasn't on the charts became ‘NA.’ These NA’s were forced to exist because of the structure of the dataset not because they are actually unknown. Therefore, we can simply ask pivot_longer to remove them by adding the argument values_drop_na = TRUE as shown below:
```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  )
#> # A tibble: 5,307 × 5
#>   artist track                   date.entered week   rank
#>   <chr>  <chr>                   <date>       <chr> <dbl>
#> 1 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk1      87
#> 2 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk2      82
#> 3 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk3      72
#> 4 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk4      77
#> 5 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk5      87
#> 6 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk6      94
#> # ℹ 5,301 more rows
```

Data is now tidy! Can further be improved to convert some of the values from strings to numbers with `mutate()` and `parse_number()`

#### Pivoting longer
Simple example using tribble which allows us to construct small tibbles by hand.
```{r}
df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)
```

we want our new (tidy) dataset to have three variables: 
id (which already exists)
measurement (the column names) 
value (the cell values)

So, we need to make pivot df longer
```{r}
df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
#> # A tibble: 6 × 3
#>   id    measurement value
#>   <chr> <chr>       <dbl>
#> 1 A     bp1           100
#> 2 A     bp2           120
#> 3 B     bp1           140
#> 4 B     bp2           115
#> 5 C     bp1           120
#> 6 C     bp2           125
```

#### Widening datasets
`pivot_wider()` allows us to handle an observation if it is scattered across multiple rows.
```{r}
cms_patient_experience
#> # A tibble: 500 × 5
#>   org_pac_id org_nm                     measure_cd   measure_title   prf_rate
#>   <chr>      <chr>                      <chr>        <chr>              <dbl>
#> 1 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_1  CAHPS for MIPS...       63
#> 2 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_2  CAHPS for MIPS...       87
#> 3 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_3  CAHPS for MIPS...       86
#> 4 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_5  CAHPS for MIPS...       57
#> 5 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_8  CAHPS for MIPS...       85
#> 6 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_12 CAHPS for MIPS...       24
#> # ℹ 494 more rows
```

The core unit being studied is an organization. But in this format, each organization is spread across six rows with one row for each measurement taken in the survey organization. We can see the complete set of values for measure_cd and measure_title by using distinct():
```{r}
cms_patient_experience |> 
  distinct(measure_cd, measure_title)
#> # A tibble: 6 × 2
#>   measure_cd   measure_title                                                 
#>   <chr>        <chr>                                                         
#> 1 CAHPS_GRP_1  CAHPS for MIPS SSM: Getting Timely Care, Appointments, and In...
#> 2 CAHPS_GRP_2  CAHPS for MIPS SSM: How Well Providers Communicate            
#> 3 CAHPS_GRP_3  CAHPS for MIPS SSM: Patient's Rating of Provider              
#> 4 CAHPS_GRP_5  CAHPS for MIPS SSM: Health Promotion and Education            
#> 5 CAHPS_GRP_8  CAHPS for MIPS SSM: Courteous and Helpful Office Staff        
#> 6 CAHPS_GRP_12 CAHPS for MIPS SSM: Stewardship of Patient Resources
```

pivot_wider() has the opposite interface to pivot_longer(): instead of choosing new column names, we need to provide the existing columns that define the values (values_from) and the column name (names_from):
```{r}
cms_patient_experience |> 
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )
#> # A tibble: 500 × 9
#>   org_pac_id org_nm                   measure_title   CAHPS_GRP_1 CAHPS_GRP_2
#>   <chr>      <chr>                    <chr>                 <dbl>       <dbl>
#> 1 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          63          NA
#> 2 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          NA          87
#> 3 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          NA          NA
#> 4 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          NA          NA
#> 5 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          NA          NA
#> 6 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          NA          NA
#> # ℹ 494 more rows
#> # ℹ 4 more variables: CAHPS_GRP_3 <dbl>, CAHPS_GRP_5 <dbl>, ...
```

Because there are still multiple rows for each organization, we also need to tell pivot_wider() which column or columns have values that uniquely identify each row; in this case those are the variables starting with "org":
```{r}
cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
#> # A tibble: 95 × 8
#>   org_pac_id org_nm           CAHPS_GRP_1 CAHPS_GRP_2 CAHPS_GRP_3 CAHPS_GRP_5
#>   <chr>      <chr>                  <dbl>       <dbl>       <dbl>       <dbl>
#> 1 0446157747 USC CARE MEDICA...          63          87          86          57
#> 2 0446162697 ASSOCIATION OF ...          59          85          83          63
#> 3 0547164295 BEAVER MEDICAL ...          49          NA          75          44
#> 4 0749333730 CAPE PHYSICIANS...          67          84          85          65
#> 5 0840104360 ALLIANCE PHYSIC...          66          87          87          64
#> 6 0840109864 REX HOSPITAL INC          73          87          84          67
#> # ℹ 89 more rows
#> # ℹ 2 more variables: CAHPS_GRP_8 <dbl>, CAHPS_GRP_12 <dbl>
```
Voila!

#### Pivoting wider
Simple example
```{r}
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)
```

Take the names from the measurement column using the `names_from()` argument and the values from the value column using the `values_from()` argument.
```{r}
df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )
#> # A tibble: 2 × 4
#>   id      bp1   bp2   bp3
#>   <chr> <dbl> <dbl> <dbl>
#> 1 A       100   120   105
#> 2 B       140   115    NA
```

To start the pivoting process, pivot_wider() needs to first figure out what will go in the rows and columns. The new column names will be the unique values of measurement.
```{r}
df |> 
  distinct(measurement) |> 
  pull()
#> [1] "bp1" "bp2" "bp3"
```

By default, the rows in the output are determined by all the variables that aren’t going into the new names or values. These are called the id_cols. Here there is only one column, but in general there can be any number.
```{r}
df |> 
  select(-measurement, -value) |> 
  distinct()
#> # A tibble: 2 × 1
#>   id   
#>   <chr>
#> 1 A    
#> 2 B
```

Combine results to generate an empty dataframe.
```{r}
df |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA)
#> # A tibble: 2 × 4
#>   id    x     y     z    
#>   <chr> <lgl> <lgl> <lgl>
#> 1 A     NA    NA    NA   
#> 2 B     NA    NA    NA
```
Then it fills in all the missing values using the data in the input. Read more about how `pivot_wider()` “makes” missing values in [chapter 19](https://r4ds.hadley.nz/missing-values#introduction) of the textbook.

EXERCISE
section 4.5.5
```{r}

```


#### Separating and uniting data tables
In table3, we see one column (rate) that contains two variables (cases and population). To address this, we can use the separate() function which separates one column into multiple columns wherever you designate.
```{r}
table3
#> # A tibble: 6 × 3
#>   country      year rate             
#> * <chr>       <int> <chr>            
#> 1 Afghanistan  1999 745/19987071     
#> 2 Afghanistan  2000 2666/20595360    
#> 3 Brazil       1999 37737/172006362  
#> 4 Brazil       2000 80488/174504898  
#> 5 China        1999 212258/1272915272
#> 6 China        2000 213766/1280428583
```

We need to split the rate column up into two variables: 1) cases and 2) population. separate() will take the name of the column we want to split and the names of the columns we want it split into. See the code below:
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"))
#> # A tibble: 6 × 4
#>   country      year cases  population
#>   <chr>       <int> <chr>  <chr>     
#> 1 Afghanistan  1999 745    19987071  
#> 2 Afghanistan  2000 2666   20595360  
#> 3 Brazil       1999 37737  172006362 
#> 4 Brazil       2000 80488  174504898 
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
```

Note from R4DS: By default, separate() will split values wherever it sees a non-alphanumeric character (i.e. a character that isn’t a number or letter). For example, in the code above, separate() split the values of rate at the forward slash characters. If you wish to use a specific character to separate a column, you can pass the character to the sep argument of separate(). For example, we could rewrite the code above as:
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")
```

Notice the data types in table3 above. Both cases and population are listed as character (<chr>) types. This is a default of using separate(). However, since the values in those columns are actually numbers, we want to ask separate() to convert them to better types using convert = TRUE. Now you can see they are listed as integer types(<int>)
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)
#> # A tibble: 6 × 4
#>   country      year  cases population
#>   <chr>       <int>  <int>      <int>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
```

Note: R4DS: You can also pass a vector of integers to sep. separate() will interpret the integers as positions to split at. Positive values start at 1 on the far-left of the strings; negative values start at -1 on the far-right of the strings. When using integers to separate strings, the length of sep should be one less than the number of names in into.

You can use this arrangement to separate the last two digits of each year. This makes this data less tidy, but is useful in other cases, as you’ll see in a little bit.
```{r}
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)
#> # A tibble: 6 × 4
#>   country     century year  rate             
#>   <chr>       <chr>   <chr> <chr>            
#> 1 Afghanistan 19      99    745/19987071     
#> 2 Afghanistan 20      00    2666/20595360    
#> 3 Brazil      19      99    37737/172006362  
#> 4 Brazil      20      00    80488/174504898  
#> 5 China       19      99    212258/1272915272
#> 6 China       20      00    213766/1280428583
```

To perform the inverse of separate() we will use unite() to combine multiple columns into a single column. In the example below for table5, we use unite() to rejoin century and year columns. unite() takes a data frame, the name of the new variable and a set of columns to combine using dplyr::select().
```{r}
table5 %>% 
  unite(new, century, year, sep = "")
#> # A tibble: 6 × 3
#>   country     new   rate             
#>   <chr>       <chr> <chr>            
#> 1 Afghanistan 1999  745/19987071     
#> 2 Afghanistan 2000  2666/20595360    
#> 3 Brazil      1999  37737/172006362  
#> 4 Brazil      2000  80488/174504898  
#> 5 China       1999  212258/1272915272
#> 6 China       2000  213766/1280428583
```
Here we need to add sep ="" because we don’t want any separator (the default is to add an underscore _)

### Handling missing values
NA is absent data, explicit absence
Blank cell is absence OF data, implicit absence

#### Explicit missing values 
