---
title: "Workshop4"
author: "Tory Pritchard"
date: "2024-05-17"
output: html_document
---

#____________________________________________________________________________

Workspace/Rmd file location
C:/Users/Admin/OneDrive/Documents/MB5370/Module_04/Mod4_Personal/Workshop4.Rmd
File also in Git hub!

Description: This workshop shows how to do GIS in R.

#____________________________________________________________________________

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

Install and load required packages
```{r}
#install.packages("sf") 
#install.packages("terra")
#install.packages("tmap")


#load into R library
library(tidyverse)
library(sf) # simple features
library (terra) # for raster
library(tmap) # Thematic maps are geographical maps in which spatial data distributions are visualized
```


### Load data
```{r}
#load the copepod data into R studio
library(readr)
dat <- read_csv("data-for-course/copepods_raw.csv")
dat
```

A tibble
```{r}
# A tibble: 5,313 × 11
   silk_id segment_no latitude longitude sample_time_utc  project route vessel  
     <dbl>      <dbl>    <dbl>     <dbl> <chr>            <chr>   <chr> <chr>   
 1       1          1    -28.3      154. 26/06/2009 22:08 AusCPR  BRSY  ANL Win...
 2       1          5    -28.7      154. 26/06/2009 23:12 AusCPR  BRSY  ANL Win...
 3       1          9    -29.0      154. 27/06/2009 0:17  AusCPR  BRSY  ANL Win...
 4       1         13    -29.3      154. 27/06/2009 1:22  AusCPR  BRSY  ANL Win...
 5       1         17    -29.7      154. 27/06/2009 2:26  AusCPR  BRSY  ANL Win...
 6       1         18    -29.8      154. 27/06/2009 2:43  AusCPR  BRSY  ANL Win...
 7       1         26    -30.4      153. 27/06/2009 4:52  AusCPR  BRSY  ANL Win...
 8       1         30    -30.7      153. 27/06/2009 5:57  AusCPR  BRSY  ANL Win...
 9       1         33    -31.0      153. 27/06/2009 6:45  AusCPR  BRSY  ANL Win...
10       1         37    -31.3      153. 27/06/2009 7:50  AusCPR  BRSY  ANL Win...
# ... with 5,303 more rows, and 3 more variables: meanlong <dbl>, region <chr>,
#   richness_raw <dbl>
```

`silk_id column`, which is just the ID for each of the silks, onto which plankton are recorded. 
For processing, silks are divided into segments, so you will also see a `segment_no` column. The other columns are pretty self explanatory.

### Check coordinates
Make first map with `ggplot2` and plot the coordinates for the samples (segments of the CPR skiils)
```{r}
ggplot(dat) + 
  aes(x = longitude, y = latitude, color = richness_raw) +
  geom_point()
```

