---
title: "MB5370 Module 4 Workshop 2 - Using ggplot2 for communication"
author: "Tory Pritchard"
date: "2024-05-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Labels

Good labels necessary for communicating findings or comply with formatting requirements for journal or company's preferred style.
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size")
```

 Add more text 

subtitle() adds additional detail in a smaller font beneath the title and caption adds text at the bottom right of the plot 
caption() adds text at the bottom right of the plot, often used to describe the source of the data

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )
```


 labs() can replace axis labels and legend titles
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    colour = "Car type"
  )
```

ggplot2 help function can tell you more things you can do with labels, including mathematical notations and superscripts.


# Annotations
Add text to the plot itself with geom_text() 
(rather than with geom_point adding shape geometry, this adds textual labels to your plots).
Allows use to pull data values out from data and include in the plot. Particularily nice when these might change on an update of the data.

```{r}
best_in_class <- mpg %>%
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), data = best_in_class)
```

These labels overlap, so use nudge() to move text a certain amount and use other R packages to handle text wrapping etc.

# Scales
Automatically added defaults by ggplot2
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))
```

Add values to the scale parameters by adding numbers in the appropriate scale arguments
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_colour_discrete()
```

Remember that to provide limits you’ll need to offer a character vector showing the start and end of your limit (e.g. c(0,12)). Note also that: ?scale_x_continuous() has plenty of other arguments, so don’t forget to explicitly state the argument you’re providing these limits for (e.g. limits = c(0,12)).

# Axis ticks
break() controls the position of the ticks and you can use labels as an argument to change the text label associated with ticks.
seq() lets you give the min and max y value and what increments you want the axis to go up by.
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5))
```

Use labels to set NULL to suppress labels altogether. Helpful when trying to format plots in a particular way.
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL)
```

# Legends and Colours
theme() used to control overall position of legend 
legend.position = "none" suppresses display of legend altogether.
```{r}
base <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))

base + theme(legend.position = "left")
base + theme(legend.position = "top")
base + theme(legend.position = "bottom")
base + theme(legend.position = "right") #default
```


# Replacing a scale
main two types to switch out: continuous position and colour scales.
ex. log transform
```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_bin2d() + 
  scale_x_log10() + 
  scale_y_log10()
```

Colour scale customization is one of the most important changes to make to a base ggplot
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv))

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  scale_colour_brewer(palette = "Set1")
```

If  there are only a few colours, add redundant shape mapping. Helps ensure plot is interpretable in black and white.
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_colour_brewer(palette = "Set1")
```

ColorBrewer scales give a range of different colour schemes you can import.
[ggplot2 cookbook](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/).
Can also use sparky colour schemes such as [Wes Anderson](https://www.r-bloggers.com/2022/07/colorful-r-plots-with-wes-anderson-palettes-pirate-ships/)

When you have predefined colours you want to use you can set them yourself, set them using scale_colour_manual()
```{r}
presidential %>%
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue"))
```

Viridis colour scheme is very popular and colour-blind friendly.
scale_colour_viridis() provided by the viridis package.
```{r}
#install.packages('viridis')
#install.packages('hexbin')
library(viridis)
library(hexbin)

df <- tibble( # note we're just making a fake dataset so we can plot it
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() + # a new geom!
  coord_fixed()

ggplot(df, aes(x, y)) +
  geom_hex() +
  viridis::scale_fill_viridis() +
  coord_fixed()
```

# Themes 
Allows you to change some or all of the non-data elements of your plot with a theme. 
ggplot2 has [8 themes by default](https://ggplot2.tidyverse.org/reference/ggtheme.html).
Many more add on packages by [Jeffrey Arnold](https://github.com/jrnold/ggthemes)
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw()
```

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_light()
```

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_classic()
```

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_dark()
```

You can develop your own themes by setting some or all of the arguments to theme()
ex. by Ben Cardwell
```{r}
 theme (panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="black"),
        axis.text.x=element_text(colour="black"),
        text=element_text(family="Arial"))
```

# Saving and exporting your plots
Best way to make a single file is exporting it using ggsave() to your working directory.
```{r}
ggplot(mpg, aes(displ, hwy)) + geom_point()

ggsave("my-plot.pdf")
#> Saving 7 x 4.32 in image
#> Use width and height arguments to change dimensions of plot.
```

# knitr
Alternative way to export results, with code, using R markdown via knitr. 

# Summary

Advance your learning by learning all the details about different geoms and all of the possible arguments you can pass to them.

Play around with customizing plots.
[Plots to look life field sites](https://cran.r-project.org/web/packages/earthtones/index.html)
Using ?colour_scale_manual to build own schemes
[Colour package generator](http://colormind.io/)
[Colour package generator](https://coolors.co/)
[Colour sampler tool](https://eyedropper.org/)
[Extra themes available](https://github.com/jrnold/ggthemes)

If you curate your own theme, you can paste a line of code at the beginning of your script to use througout
ex. (nick.theme <- theme(...))

End of workshop 2
#____________________________________________