---
title: "Workshop4"
author: "Tory Pritchard"
date: "2024-05-17"
output: html_document
---

#____________________________________________________________________________

Workspace/Rmd file location
C:/Users/Admin/OneDrive/Documents/MB5370/Module_04/Mod4_Personal/Workshop4.Rmd
File also in Git hub!

Description: This workshop shows how to do GIS in R.

#____________________________________________________________________________

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

Install and load required packages
```{r}
#install.packages("sf") 
#install.packages("terra")
#install.packages("tmap")


#load into R library
library(tidyverse)
library(sf) # simple features
library (terra) # for raster
library(tmap) # Thematic maps are geographical maps in which spatial data distributions are visualized
```


### Load data
```{r}
#load the copepod data into R studio
library(readr)
dat <- read_csv("../data/data-for-course/copepods_raw.csv")
dat
```
`silk_id column`, which is just the ID for each of the silks, onto which plankton are recorded. 
For processing, silks are divided into segments, so you will also see a `segment_no` column. The other columns are pretty self explanatory.


### Check coordinates
Make first map with `ggplot2` and plot the coordinates for the samples (segments of the CPR skiils)
```{r}
ggplot(dat) + 
  aes(x = longitude, y = latitude, color = richness_raw) +
  geom_point()
```

